Scoring Patients

### 2017-07-05 Clustering.Rmd
...

data.files = list.files(paste(data, 'QDNAseq',sep='/'), full.names=T)
plot.dir = paste(data, 'Analysis/multipcf_plots_fitted_perPatient', sep='/')

@@

## Patient info file
patient.file = grep('All_patient_info.txt', data.files, value=T)
if (length(patient.file) != 1)
  stop(paste("Missing/too many patient info file(s) in", data))

load('Training_patients.Rdata', verbose=T)

patient.info = read.patient.info(patient.file)
patient.info$Patient = gsub("/", "_", patient.info$Patient)
#head(patient.info)

@@

```{r global_pt, echo=F, message=F, warning=F, error=F}
be = subset(patient.info, Pathology == 'BE')
lgd = subset(patient.info, Pathology <= 'LGD')

@@

set.seed(99)
mtp1 = MTP(X=t(patient.info[,c('Barretts.Cellularity','Total.Reads')]), Y=as.integer((as.integer(patient.info$Pathology) > 3)), 
           get.adjp=T, na.rm=T, test='t.twosamp.equalvar')
mtp2 = MTP(X=t(patient.info[,c('Barretts.Cellularity','Total.Reads')]), Y=patient.info$Status, 
            get.adjp=T, na.rm=T, test='t.twosamp.equalvar')
```

Currently with multi test correction the samples for which cellularity and #of reads are available show no difference for either status (P vs NP)
`r pander(mtp2@adjp, justify='left', caption="Adjusted p-value")`

or for samples that are LGD/IMC/HGD vs BE
`r pander(mtp1@adjp, justify='left', caption="Adjusted p-value")`

However, it should be noted that pathologists do not like calling cellularity on Barrett's generally so it's a fairly inaccurate measure.
`r pander(table( patient.info[,c('Status','Barretts.Cellularity')]), caption="Numbers of cellularity calls by status")`

With regards to p53 staining, non-progressors have no p53 positive tests, while progressors had `r p53Status[2,2]` overall.
`r pander(p53Status, justify='left', caption="P53 staining status for progressors vs non")`
Of these `r table(be$p53.Status)[['1']]` were positive in a BE sample, and `r table(lgd$p53.Status)[['1']]` were positive when LGD samples were included.

@@

apclust.data<-function(segdata, samples) {
  x1 = segdata[,intersect(colnames(segdata), samples)]
  #x1 = x1[, samples]
  rownames(x1) = (segdata[,c(1:4)] %>%
    rowwise() %>%
    mutate(location=paste(paste(chrom, arm, sep=''), '.', start.pos, '-', end.pos, sep='')))$location
  return(x1)  
}

@@

## Using the segmentation values
sum.patient.data$years = with(sum.patient.data, paste(start.year, end.year, sep='-'))
sim = as.data.frame(sum.patient.data[,c('Status','years')])
rownames(sim) = sum.patient.data$Patient
for (pt in sum.patient.data$Patient) {
  patient.data[[pt]]$info = arrange(patient.data[[pt]]$info,  Pathology, Endoscopy.Year)
  x1 = apclust.data(patient.data[[pt]]$seg.vals, patient.data[[pt]]$info$Samplename)

  gg = ggplot(melt(x1), aes(x=variable, y=value, group=variable, fill=variable)) + 
    geom_boxplot(notch=T, show.legend=F) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position='none') +
    labs(title=paste(pt, patient.data[[pt]]$info$Status[1], sep=' - ')) +   theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

@@

  sim[pt, 'ns'] = net.similarity(ac)
  sim[pt, 'ss'] = sum.similarity(ac)
  sim[pt, 'clusters'] = length(clusters(ac))
  sim[pt, 'max.sd'] = max(sapply(x1, sd))

  tmp = patient.data[[pt]]$variable.regions[,-(1:5)]
  sim[pt, 'n.var'] = nrow(patient.data[[pt]]$variable.regions[,-(1:5)])

  if (nrow(tmp) <= 0) {
    sim[pt,c('gains', 'loss')] = c(0,0)
  } else {
    sim[pt,'gains'] = length(tmp[tmp>=1.1])
    sim[pt,'loss'] =  length(tmp[tmp<=0.9])
  }
}
sim$id = rownames(sim)
rc = cor.test(sim$max.sd, sim$clusters)

do.call(grid.arrange, c(sample.vals))

@@

```{r apclustN, echo=F, message=F, warning=F, fig.height=5, fig.width=5, fig.align='left'}
# Should this be performed on the normalized seg values instead?
sim3 = sum.patient.data[,c(2:4)]
rownames(sim3) = sum.patient.data[,1]
for (pt in sum.patient.data$Patient) {
  x1 = patient.data[[pt]]$norm.seg.vals
  ac = apcluster(negDistMat(x1, r=2), q=q, details=T, includeSim=T)
  #plot(ac)

>>>>

data.files = list.files(paste(data, 'QDNAseq',sep='/'), full.names=T)
analysis.files = list.files(paste(data, 'Analysis', sep='/'), full.names=T)
plot.dir = paste(data, 'Analysis/multipcf_plots_fitted_perPatient', sep='/')

@@

## Patient info file
patient.file = grep('All_patient_info.txt', data.files, value=T)
if (length(patient.file) != 1)
  stop(paste("Missing/too many patient info file(s) in", data))
demo.file = grep('Demographics_full.txt', data.files, value=T)

load(grep('Training_patients.Rdata', analysis.files, value=T), verbose=T)

patient.info = read.patient.info(patient.file, demo.file)
#head(patient.info)

@@

```{r global_pt, echo=F, message=F, warning=F, error=F, eval=F}
be = subset(patient.info, Pathology == 'BE')
lgd = subset(patient.info, Pathology <= 'LGD')

@@

set.seed(99)
mtp1 = MTP(X=t(patient.info[,c('Barretts.Cellularity','Total.Reads')]), Y=as.integer((as.integer(patient.info$Pathology) > 3)), 
           get.adjp=T, na.rm=T, test='t.twosamp.equalvar')
mtp2 = MTP(X=t(patient.info[,c('Barretts.Cellularity','Total.Reads')]), Y=patient.info$Status, 
            get.adjp=T, na.rm=T, test='t.twosamp.equalvar')

# Currently with multi test correction the samples for which cellularity and #of reads are available show no difference for either status (P vs NP)
# `r pander(mtp2@adjp, justify='left', caption="Adjusted p-value")`
# 
# or for samples that are LGD/IMC/HGD vs BE
# `r pander(mtp1@adjp, justify='left', caption="Adjusted p-value")`
# 
# However, it should be noted that pathologists do not like calling cellularity on Barrett's generally so it's a fairly inaccurate measure.
# `r pander(table( patient.info[,c('Status','Barretts.Cellularity')]), caption="Numbers of cellularity calls by status")`
# 
# With regards to p53 staining, non-progressors have no p53 positive tests, while progressors had `r p53Status[2,2]` overall.
# `r pander(p53Status, justify='left', caption="P53 staining status for progressors vs non")`
# Of these `r table(be$p53.Status)[['1']]` were positive in a BE sample, and `r table(lgd$p53.Status)[['1']]` were positive when LGD samples were included.

@@

apclust.data<-function(segdata, samples) {
  x1 = as.matrix(segdata[,intersect(colnames(segdata), samples)])
  #x1 = x1[, samples]
  rownames(x1) = (segdata[,c(1:4)] %>%
    rowwise() %>%
    mutate(location=paste(paste(chrom, arm, sep=''), '.', start.pos, '-', end.pos, sep='')))$location
  if (length(samples) == 1) colnames(x1) = samples
  return(x1)  
}

@@

## Using the segmentation values
sum.patient.data$years = with(sum.patient.data, paste(start.year, end.year, sep='-'))
sim = as.data.frame(sum.patient.data[,c('Status','years')])
rownames(sim) = sum.patient.data$Patient
for (pt in sum.patient.data$Patient) {
  print(pt)
  patient.data[[pt]]$info = arrange(patient.data[[pt]]$info,  Pathology, Endoscopy.Year)
  x1 = apclust.data(patient.data[[pt]]$seg.vals, patient.data[[pt]]$info$Samplename)

  gg = ggplot(melt(x1), aes(x=Var2, y=value, group=Var2, fill=Var2)) + 
    geom_boxplot(notch=T, show.legend=F) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position='none') +
    labs(title=paste(pt, patient.data[[pt]]$info$Status[1], sep=' - '), x='')

@@

  sim[pt, 'ns'] = net.similarity(ac)
  sim[pt, 'ss'] = sum.similarity(ac)
  sim[pt, 'clusters'] = length(clusters(ac))
  sim[pt, 'max.sd'] = max(apply(x1, 2, sd))

  tmp = patient.data[[pt]]$variable.regions[,-(1:5)]
  sim[pt, 'n.var'] = ifelse(!is.null(tmp) && nrow(tmp)>0, nrow(patient.data[[pt]]$variable.regions[,-(1:5)]), 0)

  if (is.null(tmp) || nrow(tmp) <= 0) {
    sim[pt,c('gains', 'loss')] = c(0,0)
  } else {
    sim[pt,'gains'] = length(tmp[tmp>=1.1])
    sim[pt,'loss'] =  length(tmp[tmp<=0.9])
  }
}
sim$id = rownames(sim)
rc = cor.test(sim$max.sd, sim$clusters)

#do.call(grid.arrange, c(sample.vals))

@@

```{r apclustN, echo=F, message=F, warning=F, fig.height=5, fig.width=5, fig.align='left'}
# Should this be performed on the normalized seg values instead?
sim3 = sum.patient.data[,c(2:4)]
rownames(sim3) = sum.patient.data[,1]
for (pt in sum.patient.data$Patient) {
  print(pt)
  x1 = patient.data[[pt]]$norm.seg.vals
  if (is.null(x1)) next
  ac = apcluster(negDistMat(x1, r=2), q=q, details=T, includeSim=T)
  #plot(ac)

### 2017-07-05 Patient Report.Rmd
More analysis
...

---
title: "Scoring Patients"
author: "Sarah Killcoyne"
date: "23 May 2017"
output:
  html_document: 
    toc: yes
  word_document: default
---

@@

## Patient info file
patient.file = grep('/All_patient_info.txt', data.files, value=T)
if (length(patient.file) != 1)
  stop(paste("Missing/too many patient info file(s) in", data))

dataset = 'Training'
patient.info = read.patient.info(patient.file, set=dataset)
head(patient.info)

@@

n.var.regions = sapply(var.seg.lengths, nrow)
wilcox.test(n.var.regions[prog.pt], n.var.regions[np.pt])

@@

@@

```{r, hmm, echo=F, message=F, warning=F, fig.height=100, fig.width=10}
ccgenes = read.table('~/Data/CosmicCensusGenes.tsv', sep='\t', header=T, stringsAsFactors=F)

noLoc = grep(':-',ccgenes$Genome.Location)
noLocGenes = ccgenes[noLoc,]

ccgenes = ccgenes[-noLoc,]

@@

gene.cnt = arrange(gene.cnt, -Progressors)
## RAD51B is screwing up the plot simply because it shows 16 non-progressors (8 progressors) with a gain but the next closest is 9 so we'll just filter that out for now
gene.cnt = subset(gene.cnt, gene.cnt$Gene.Symbols != 'RAD51B')

@@

m$brks <- cut(m$value, include.lowest=T, right=F,ordered_result=T,
                   breaks=c(0, 1, 3, 5, 8, 10),
                   labels=c('0','1-3', '4-5', '6-8','9-10'))

@@

m$brks <- cut(m$value, include.lowest=T, right=F,
                   breaks=c(0, 1, 3, 5, 8, 10),
                   labels=c('0','1-3', '4-5', '6-8','9-10'))

>>>>

---
title: "Scoring Patients"
author: "Sarah Killcoyne"
date: "22 June 2017"
output:
  html_document: 
    toc: yes
  word_document: default
---

@@

## Patient info file
patient.file = grep('/All_patient_info.txt', data.files, value=T)
if (length(patient.file) != 1)
  stop(paste("Missing/too many patient info file(s) in", data))
demo.file = grep('/Demographics_full.txt', data.files, value=T)

dataset = 'Test'
patient.info = read.patient.info(patient.file, demo.file, set=dataset)
head(patient.info)

@@

sum.var = rbind('Progressors'=do.call(rbind.data.frame, var.seg.lengths[prog.pt]) %>% summarize( 'n.segments' = length(seg.len), 'median.length'=median(seg.len), 'mean.length'=mean(seg.len), 'sd'=sd(seg.len)),
  'Non-Progressors'=do.call(rbind.data.frame, var.seg.lengths[np.pt]) %>% summarize( 'n.segments' = length(seg.len), 'median.length'=median(seg.len), 'mean.length'=mean(seg.len), 'sd'=sd(seg.len)))

pander(sum.var, caption='Variable regions between progressors and non-progressors')

sum.seg = rbind('Progressors'=do.call(rbind.data.frame, segment.lengths[prog.pt]) %>% summarize( 'n.segments' = length(seg.len), 'median.length'=median(seg.len), 'mean.length'=mean(seg.len), 'sd'=sd(seg.len)),
  'Non-Progressors'=do.call(rbind.data.frame, segment.lengths[np.pt]) %>% summarize( 'n.segments' = length(seg.len), 'median.length'=median(seg.len), 'mean.length'=mean(seg.len), 'sd'=sd(seg.len)))

pander(sum.var, caption='Segment stats between progressors and non-progressors')
```

```{r echo=F, message=F, warning=F}

df = rbind(cbind(do.call(rbind.data.frame, var.seg.lengths[prog.pt]), 'status'='Progressor'), cbind(do.call(rbind.data.frame, var.seg.lengths[np.pt]), 'status'='Non-Progressor'))

lenWT = wilcox.test(subset(df, status == 'Progressor')$seg.len, subset(df, status == 'Non-Progressor')$seg.len)

ggplot(df, aes(status, seg.len)) + 
  geom_boxplot(aes(group=status, fill=status), show.legend=F) + 
  geom_label(data=subset(df, status == 'Progressor'), aes(status, max(seg.len)-2*sd(seg.len), label=paste('median\n',median(seg.len),sep='') )) +
  geom_label(data=subset(df, status == 'Non-Progressor'), aes(status, max(seg.len), label=paste('median\n',median(seg.len),sep=''))) +
  labs(y='Variable segment length', x='', title='Variable regions', caption=paste('Wilcox p-value', ifelse (lenWT$p.value < .001, '< .001', format.pval(lenWT$p.value)) ))

df = rbind(cbind.data.frame('n'=sapply(var.seg.lengths[prog.pt], nrow), 'status'='Progressor'), cbind.data.frame('n'=sapply(var.seg.lengths[np.pt], nrow), 'status'='Non-Progressor'))

nWT = wilcox.test(subset(df, status == 'Progressor')$n, subset(df, status == 'Non-Progressor')$n)

ggplot(df, aes(status, n, group=status)) + 
  geom_boxplot(aes(fill=status), position = 'dodge', show.legend = F) + 
  geom_label(data=subset(df, status == 'Progressor'), aes(status, max(n)-2*sd(n), label=paste('median\n',median(n),sep='') )) +
  geom_label(data=subset(df, status == 'Non-Progressor'), aes(status, max(n), label=paste('median\n',median(n),sep=''))) +
  labs(y='Number of segments', title="Total number of segments", caption=paste('Wilcox p-value', ifelse (lenWT$p.value < .001, '< .001', format.pval(nWT$p.value)) ))

### Check variable segment length vs chromosome length

```{r echo=F, message=F, warning=F}
n.var.regions = sapply(var.seg.lengths, nrow)

wilcox.test(n.var.regions[prog.pt], n.var.regions[np.pt])

@@

## Gain vs Loss

```{r echo=F, message=F, warning=F}

df = do.call(rbind.data.frame, lapply(patient.data, function(df) {
  nohgd = subset(df$info, !Pathology %in% c('HGD','IMC'))$Samplename
  gain=0;loss=0
  if ( !is.null(df[['variable.regions']]) ) {
    nohgd = intersect(nohgd, colnames(df$variable.regions))
    gain = length(which(apply(as.data.frame(df$variable.regions[,nohgd]), 1, function(x) length(which(x >= 1.1))) > 0))
    loss = length(which(apply(as.data.frame(df$variable.regions[,nohgd]), 1, function(x) length(which(x <= 0.9))) > 0))
  }

  cbind.data.frame('gain'=gain, 'loss'=loss, 'status'=unique(df$info$Status))
}))
head(df)

ttg = t.test(subset(df, status == 'P')$gain, subset(df, status == 'NP')$gain)
ttl = t.test(subset(df, status == 'P')$loss, subset(df, status == 'NP')$loss)

ft = fisher.test((df %>% group_by(status) %>% summarize( 'gains'=sum(gain), 'losses'=sum(loss)))[,c('gains','losses')])

ggplot(melt(df), aes(status, value)) + facet_grid(.~variable) + 
  geom_bar(aes(fill=status),stat='identity') + 
  labs(title='Total gains & Losses excluding HGD/IMC', y='counts', x='', caption=paste('Gains and losses are signficantly enriched in Progressors vs Non. Fisher odds ', round(ft$estimate,2), '(p=', format.pval(ft$p.value, digits=1), ')', sep='' ))

@@

```{r, hmm, echo=F, message=F, warning=F, fig.height=100, fig.width=10}
ccgenes = read.table('~/Data/CosmicCensusGenes.tsv', sep='\t', header=T, stringsAsFactors=F)
ccgenes = ccgenes[-which(ccgenes$Mutation.Types == 'T'),]

noLoc = grep(':-',ccgenes$Genome.Location)
if (length(noLoc) > 0) {
  noLocGenes = ccgenes[noLoc,]
  ccgenes = ccgenes[-noLoc,]
}

@@

gene.cnt = arrange(gene.cnt, -Progressors)
## RAD51B is screwing up the plot simply because it shows 16 non-progressors (8 progressors) with a gain but the next closest is 9 so we'll just filter that out for now
#gene.cnt = subset(gene.cnt, gene.cnt$Gene.Symbols != 'RAD51B')

@@

m$brks <- cut(m$value, include.lowest=T, right=F,ordered_result=T,
                   breaks=c(0, 1, 3, 5, 8, 10, 12),
                   labels=c('0','1-3', '4-5', '6-8','9-10', '10-12'))

@@

m$brks <- cut(m$value, include.lowest=T, right=F,
                   breaks=c(0, 1, 3, 5, 8, 10, 12),
                   labels=c('0','1-3', '4-5', '6-8','9-10', '10-12'))
